Measurement of a dwarf galaxy rotational velocity from its Lyman-alpha line emission

### 2016-09-22 paper/letter.tex
referencias, nuevo texto

\newcommand{\tol}{Tololo 1214-277}
\newcommand{\lya}{Ly$\alpha$}
\newcommand{\hb}{H$\beta$}
\newcommand{\ha}{H$\alpha$}
\newcommand{\oiii}{[OIII]}
\newcommand{\oii}{[OII]}
\newcommand{\nii}{[NII]}
\newcommand{\esca}{erg cm$^{-2}$ s$^{-1}$ \AA$^{-1}$}
\newcommand{\esc}{erg cm$^{-2}$ s$^{-1}$}
\newcommand{\es}{erg s$^{-1}$}
\newcommand{\esa}{erg s$^{-1}$}
\newcommand{\kms}{km s$^{-1}$}
\newcommand{\apj}{ApJ}  
\newcommand{\jcap}{JCAP}  
\newcommand{\apjs}{ApJS}  
\newcommand{\pasa}{PASA}  
\newcommand{\apjl}{ApJL}  
\newcommand{\aj}{AJ}  
\newcommand{\mnras}{MNRAS}  
\newcommand{\mnrassub}{MNRAS accepted}  
\newcommand{\aap}{A\&A}  
\newcommand{\aaps}{A\&AS}  
\newcommand{\araa}{ARA\&A}  
\newcommand{\nat}{Nature}  
\newcommand{\physrep}{PhR}
\newcommand{\pasp}{PASP}    
\newcommand{\pasj}{PASJ}    
\newcommand{\sigmaclump}{$69^{+17}_{-15}$ km s$^{-1}$}
\newcommand{\inftyclump}{$79^{+73}_{-42}$ km s$^{-1}$}
\newcommand{\probaclump}{$0.78^{+0.16}_{-0.43}$}
\def\simgt{\lower.5ex\hbox{\gtsima}}
\def\simlt{\lower.5ex\hbox{\ltsima}}
%------------------------------------------------------

\begin{document}
\pagestyle{empty}
\noindent
\textbf{Line sketches of a massive dwarf galaxy}
\\
\\
Jaime E. Forero-Romero$^{1}$, Max Groenke$^2$, Mar\'ia Camila
Remolina-Guti\'errez$^1$, Juan Nicol\'as
Garavito-Carmargo$^3$, Mark Dijkstra$^2$.
\\
\\
\scriptsize
{$^1$ Departamento de Física, Universidad de los Andes, Cra. 1
  No. 18A-10 Edificio Ip, CP 111711, Bogot\'a, Colombia 
\\
$^2$ Institute of Theoretical Astrophysics, University of Oslo,
Postboks 1029 Blindern, NO-0315 Oslo, Norway.
\\
$^3$ Department of Astronomy, University of Arizona, 933 North Cherry
Avenue, Tucson, AZ 85721, USA. 
\normalsize
\\
\\
\textbf{
  Star-forming Compact Dwarf Galaxies (CDGs) resemble the expected
  pristine conditions of the first galaxies in the Universe.    
Before the observational detection of the first galaxies becomes
reality, CDGs are the best systems to test our ideas on primordial
galaxy formation and evolution.    
Here we report on one of such CDGs, \tol, which presents
a broad symmetric Lyman-$\alpha$ line emission that had evaded theoretical
interpretation so far. 
In this letter we explain these features by two different models: an
homogeneous gaseous sphere undergoing bulk rotation and an interstellar
medium composed by outflowing clumps with additional random motions.
It is the first time that an observed \lya\ spectrum can be explained
assuming either of these physical conditions.
We find that both models independently require high velocities
(either a bulk rotation of $348^{+75}_{-48}$ km s$^{-1}$ or a clump velocity
dispersion of \sigmaclump with outflows of
$79^{+73}_{-42}$ km s$^{-1}$) consistent with a dynamical mass at
least $10$ times larger than its baryonic mass.  
We argue that the most plausible explanation for this excess of
dynamical mass is the presence of a supermassive black hole at the
center of \tol. 
This work demonstrates the importance of considering multiphase
physics and rotation among the possible conditions shaping the
\lya\ spectra of the first galaxies.  
Additionally, if future kinematic maps of \tol\ confirm the high
velocities postulated in our model, it would provide new
evidence for dwarf galaxies as hosts of supermassive black
holes.  
}

@@

\tol\ is a compact star forming dwarf galaxy that presents a
strong Ly-$\alpha$ emission \cite{Thuan97} with two puzzling 
features: the line is symmetric and single peaked.
Usually the Ly-$\alpha$ line has a single or asymmetric double peak. 
These two special features in \tol\ cannot be explained with
conventional models \cite{2006A&A...460..397V,2015ApJ...812..123G}.

In this letter we show how the \tol's \lya\ profile can be explained
either by rotation or the recently developed class of more complex
multiphase models that predict a wider variety of spectra
including, single, double and triply peaked spectra.  
Figure 1. summarizes our findings.
Dots represent the observational data for \tol\ with the
overplot from our best fits from the analytical solution for a
rotating homogeneous gas sphere (continuous line) and the multiphase
model (dashed line). 
This is the first time that these models have been introduced with
success to explain an observed \lya\ profile.

@@

In the multiphase model the best constrained parameters by the
observational data are the clump velocity dispersion
$\sigma_{\rm{cl}}=$\sigmaclump\,  the clump's outflowing velocity
$v_{\infty, \rm{cl}}=$\inftyclump\ and the fraction of the
\lya\ emission that is  coming from the cold clumps  $P_{\rm cl}=$\probaclump.

Assuming that the clumps are located in a spherical region of radius
$r_s\approx 2.0$ kpc (corresponding to the estimated 3D half-light
radius), 
the results from the rotational velocity correspond to a
dynamical mass of $M_{\rm dyn} \approx 5.45\times 10^{10} M_{\odot}$,
while the velocity dispersion in the multiphase model corresponds to a
$M_{\rm dyn}=2.8^{+1.3}_{-1.2}\times 10^{9}$ $M_{\odot}$.

Are these values atypical? If \tol\ followed the fundamental
plane relationship between its mean surface brightness
$I_e$, the projected half-light radius $R_e$ and the velocity dispersion
$\sigma$, described by $\log I_e=1.6 \log\sigma - 1.21\log R_e +
0.55$ \cite{2009ApJ...698.1590G},  the expected velocity dispersion
should be on the order of $5 \pm 1$ 
\kms, which is a factor of $\sim$ $12$ - $60$ lower than the 
results from the multiphase and rotation models, respectively. 
These are equivalent to factors of $\sim$ $140$ - $3600$ on the
dynamical mass;
\tol\ seems to be significantly more massive than the expectation from
scaling relationships for dwarf ellipticals.
A factor of $100$ in dynamical mass could be accounted by dark matter if
\tol\ was hosted by a dark matter halo of  $\sim 10^{16}$ M$_{\odot}$
\cite{2011ApJ...726..108T}, which is very unlikely as it would mean that
\tol\ is at the center of a super rich cluster of galaxies,
contradicting observational evidence.
The remaining possibility is that \tol\ hosts a supermassive black hole of
$\sim 5\times 10^{9}$ M$_{\odot}$, corresponding to $10$ times its
estimated stellar and gaseous mass.

@@

The optical emission  comes from a   region with approximate diameter
4 kpc \cite{Fricke01}.

@@

@@

The rotation model corresponds to the work presented in
\cite{GaravitoCamargo2014}. 
In that model the Ly-$\alpha$ photons are propagated 
within a spherical and homogeneous cloud of HI gas undergoing solid
body rotation.
The sphere is fully characterized by three parameters: the optical
depth $\tau$ measured from the center to its surface, the HI
temperature $T$, and the linear surface velocity $V_{\rm max}$.  
Photons are emitted at their natural frequency from the center of the
sphere. 
Including the effect of dust only changes the overall line
normalization but not its shape.  
The results we report in the main body of the paper do not include any
dust model.
In the current work, the radiative transfer simulations were done
using Monte Carlo simulations with the code \texttt{CLARA}
\cite{CLARA}.

@@

\section*{The Outflow Model}
The galaxy outflow model is a highly successful framework to fit
and interpret \lya\ line observations.

>>>>

\newcommand{\tol}{Tololo 1214-277}
\newcommand{\lya}{Ly$\alpha$}
\newcommand{\hb}{H$\beta$}
\newcommand{\ha}{H$\alpha$}
\newcommand{\oiii}{[OIII]}
\newcommand{\oii}{[OII]}
\newcommand{\nii}{[NII]}
\newcommand{\esca}{erg cm$^{-2}$ s$^{-1}$ \AA$^{-1}$}
\newcommand{\esc}{erg cm$^{-2}$ s$^{-1}$}
\newcommand{\es}{erg s$^{-1}$}
\newcommand{\esa}{erg s$^{-1}$}
\newcommand{\kms}{km s$^{-1}$}
\newcommand{\apj}{ApJ}  
\newcommand{\jcap}{JCAP}  
\newcommand{\apjs}{ApJS}  
\newcommand{\pasa}{PASA}  
\newcommand{\apjl}{ApJL}  
\newcommand{\aj}{AJ}  
\newcommand{\mnras}{MNRAS}  
\newcommand{\mnrassub}{MNRAS accepted}  
\newcommand{\aap}{A\&A}  
\newcommand{\aaps}{A\&AS}  
\newcommand{\araa}{ARA\&A}  
\newcommand{\nat}{Nature}  
\newcommand{\physrep}{PhR}
\newcommand{\pasp}{PASP}    
\newcommand{\pasj}{PASJ}    
\newcommand{\sigmaclump}{$54.3\pm 0.6$ km s$^{-1}$}
\newcommand{\inftyclump}{$54.3\pm 5.1$ km s$^{-1}$}
\newcommand{\probaclump}{$0.96\pm 0.01$}
\def\simgt{\lower.5ex\hbox{\gtsima}}
\def\simlt{\lower.5ex\hbox{\ltsima}}
%------------------------------------------------------

\begin{document}
\pagestyle{empty}
\noindent
\textbf{Line Sketches of a Massive Dwarf Galaxy}
\\
\\
Jaime E. Forero-Romero$^{1}$, Max Groenke$^2$, Mar\'ia Camila
Remolina-Guti\'errez$^1$, Juan Nicol\'as
Garavito-Carmargo$^3$, Mark Dijkstra$^2$.
\\
\\
\scriptsize
{$^1$ Departamento de Física, Universidad de los Andes, Cra. 1
  No. 18A-10 Edificio Ip, CP 111711, Bogot\'a, Colombia 
\\
$^2$ Institute of Theoretical Astrophysics, University of Oslo,
Postboks 1029 Blindern, NO-0315 Oslo, Norway.
\\
$^3$ Department of Astronomy, University of Arizona, 933 North Cherry
Avenue, Tucson, AZ 85721, USA. 
\normalsize
\\
\\
\textbf{
  Star-forming Compact Dwarf Galaxies (CDGs) resemble the expected
  pristine conditions of the first galaxies in the Universe.    
Before the observational detection of the first galaxies becomes
reality, CDGs are the best systems to test our ideas on primordial
galaxy formation and evolution.    
Here we report on one of such CDGs, \tol, which presents
a broad symmetric Lyman-$\alpha$ line emission that had evaded theoretical
interpretation so far. 
In this letter we explain these features by two different models: an
homogeneous gaseous sphere undergoing bulk rotation and an interstellar
medium composed by outflowing clumps with additional random motions.
It is the first time that an observed \lya\ spectrum can be explained
assuming either of these physical conditions.
We find that both models independently require high velocities
(either a bulk rotation of $348^{+75}_{-48}$ km s$^{-1}$ or a clump velocity
dispersion of \sigmaclump with outflows of
\inftyclump) consistent with a dynamical mass of at
least a billion solar masses, $7$ times larger than its baryonic mass.   
We argue that the most plausible explanation for this excess of
dynamical mass is the presence of a supermassive black hole at the
center of \tol. 
This work demonstrates the importance of considering multiphase
physics and rotation among the possible conditions shaping the
\lya\ spectra of the first galaxies.  
Additionally, if future kinematic maps of \tol\ confirm the high
velocities postulated in our model, it would provide new
evidence for dwarf galaxies as hosts of supermassive black
holes.  
}

@@

\tol\ is a compact star forming dwarf galaxy that presents a
strong Ly-$\alpha$ emission \cite{Thuan97} with two puzzling 
features: the line is symmetric and single peaked.
Usually the Ly-$\alpha$ line has an asymmetric single or double peak. 
These two special features in \tol\ cannot be explained with
conventional models \cite{2006A&A...460..397V,2015ApJ...812..123G}.

In this letter we show how the \tol's \lya\ profile can be explained
either by rotation\cite{GaravitoCamargo2014}or the recently developed
class of more complex 
multiphase models that predict a wider variety of spectra
including, single, double and triply peaked spectra \cite{Gronke2016}.  
Figure 1. summarizes our findings.
Dots represent the observational data for \tol\ with the
overplot from our best fits from the analytical solution for a
rotating homogeneous gas sphere (thin line) and the multiphase
model (thick line). 
This is the first time that these models have been introduced with
success to explain an observed \lya\ profile.

@@

In the multiphase model the best constrained parameters by the
observational data are the clump velocity dispersion
$\sigma_{\rm{cl}}=$\sigmaclump ,  the clump's outflowing velocity
$v_{\infty, \rm{cl}}=$\inftyclump\ and the fraction of the
\lya\ emission that is  coming from the cold clumps  $P_{\rm cl}=$\probaclump.

Assuming that the clumps are located in a spherical region of radius
$r_s=2.0$ kpc (corresponding to \tol's estimated 3D half-light
radius), 
this corresponds to dynamical masses of $M_{\rm dyn} =
2.8_{-0.8}^{+1.5}\times 10^{10} M_{\odot}$ and $M_{\rm
  dyn}=1.71\pm0.04 \times 10^{9}$ $M_{\odot}$ for the
rotation and multiphase models, respectively,

\tol's stellar mass is  $M_{\star} = 1.45\pm0.45\times 10^{8}
M_{\odot}$   \cite{2014PASP..126.1079M} and its total neutral HI mass is $M_{\rm HI}<2.65\times 10^{8}$ M$_{\odot}$
\cite{pustilnikmartin07}; the dynamical mass is at
least 7 to 68 times the baryonic mass, depending if one
considers the multiphase or rotation estimate.

A dynamical mass of $10^{9} M_{\odot}$ in a sphere of 2 kpc could be
accomodated by a dark matter halo of at least $10^{13}$
M$_{\odot}$ in mass \cite{2011ApJ...726..108T}, which
seems unlikely as it would mean that  \tol\ is at the center of a
galaxy group.
The remaining possibility is that \tol\ hosts a supermassive black hole of
at least $10^{9} M_{\odot}$.

Another perspective to appreciate the atypically high dynamical mass
estimates comes from the observed scaling relations for dwarf
galaxies.
Assuming that \tol\ followed the fundamental plane relationship
between its mean surface brightness $I_e$, the projected half-light
radius $R_e$ and the velocity dispersion $\sigma$, described by $\log
I_e=1.6 \log\sigma - 1.21\log R_e + 0.55$ \cite{2009ApJ...698.1590G},
the expected velocity dispersion should be on the order of $5 \pm 1$ 
\kms, which is a factor of $\sim$ $10$ - $60$ lower than the 
results from the multiphase and rotation models, respectively. 
These are equivalent to factors of $\sim$ $100$ - $3600$ on the
dynamical mass.
Once again,  \tol\ seems to be significantly more massive than
expected.

@@

@@

We computed the projected half-luminosity radius to be $R_s=1.5\pm0.1$ kpc 
from the surface intensity profiles reported by.
\cite{2003A&A...410..481N}. 
Assuming spherical geometry, one can translate this value into a 3D
half-luminosity radius of $r_s=3R_s/2=2.0$ kpc.

@@

The rotation model corresponds to the work presented in
\cite{GaravitoCamargo2014} based on the Monte Carlo code
\texttt{CLARA} \cite{CLARA}. 
In that model the Ly-$\alpha$ photons are propagated 
within a spherical and homogeneous cloud of HI gas undergoing solid
body rotation.
The sphere is fully characterized by three parameters: the HI optical
depth $\tau$ measured from the center to its surface, the HI
temperature $T$, and the linear surface velocity $V_{\rm max}$.  
Photons are emitted at their natural frequency from the center of the
sphere. 
Including the effect of dust only changes the overall line
normalization but not its shape.  
The results we report in the main body of the paper do not include any
dust model.
In this letter we use an analytical solution that captures the most
important effects of rotation onto the \lya\ line.

@@
